---
title: "How to calculate marketing impact when you can't A/B test"
subtitle: "An example of before-and-after analysis in R"
author: "Anita Owens"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
#TOC {
  color: #000000;
  font-family: Calibri;
  font-size: 16px; 
  border-color: #DA4253;
}
#header {
  color: #000000;
  background-color: #DA4253; <! -- red color -->
  opacity: 1;
  font-family: Calibri;
  font-size: 20px;
  color: #FFFFFF;
}
body {
  color: #000000;
  font-family: Calibri;
  background-color: #ffffff;
  font-size: 18px;
}
pre {
  color: #000000;
  background-color: #ffffff;
}
</style>
```
## 1. Introduction

How to calculate marketing impact when you can't A/B test.



One of the key activities for companies to become more data-driven is
A/B testing. There are going to be times where running a control
experiment for something like activating a new channel, may not be
possible. Also, customer behavior can sometimes be difficult to untangle amongst all of the different factors that influence buying decisions.

So how can we evaluate the impact of a change that's already taken place? We can use a before and after analysis. Let's get started. Rather than using a pre-selected dataset, we will create a synthetic dataset to use for our analysis.

```{r Set up environment and load packages}

# Install pacman if needed
if (!require("pacman")) install.packages("pacman")

# load packages
pacman::p_load(pacman,
  tidyverse, openxlsx, psych, ggcorrplot, corrplot, cowplot, ggthemes)
```

## 2. Create synthetic data

First, set the seed for reproducibility when simulating data especially
if you want to re-create the same dataset later.

```{r Set seed for reproducibility}
#Set seed for reproducibility
set.seed(32250)
```

The dataset we want to create is as follows:

1.  Sales date (based on week level granularity)
2.  Revenue
3.  Paid social ads

```{r Create a sequence of dates}
sales_date <- seq(as.Date('2021-01-01'), by = 'week', length.out=52)

#Check results
print(sales_date)
```

```{r Create temporary social vectors}

social_ad_spend_avg <- 220

social_1 <-rep(0, length(sales_date) - 20)

print(social_1)


#Represents 10 weeks of social ad spending
social_2 <-rep(social_ad_spend_avg, length(sales_date) - length(social_1) - 10)

print(social_2)

social_3 <-rep(0, length(sales_date) - length(social_1) - length(social_2))

print(social_3)
```

```{r Combine the 3 temporary social vectors into 1}
#Combine into 1 vector
social <- c(social_1, social_2, social_3)

#Check if we have 52 weeks worth of data
length(social)
```

Next is to generate revenue and place into a temporary revenue variable.

Revenue data is randomly generated based on a normal distribution.

```{r Generate Revenue Data}
#first argument is the number of draws
#2nd argument is the mean value per week of sales
temp_revenue <- rnorm(n=length(sales_date), mean = 3800, sd = 500)


#Check output
temp_revenue
```

```{r Create dataframe by merging all of our lists }
#Initialize data frame
df <- data.frame(sales_date,
                 temp_revenue, social)
```

In our last step, we want to scale revenue up for the periods during and after additional marketing ad spend.



```{r We want to scale sales up for the period during and after additional marketing ad spend}
df1 <- df %>% 
    mutate(revenue =
      case_when(
        sales_date >= "2021-08-13" ~ (temp_revenue + (temp_revenue * 0.35)), 
        TRUE ~ temp_revenue)
      )

head(df1)
```

The first thing that we want to do is define our (before and after) time periods, we need to add a column for the advertising periods.


```{r Create new variable for before and after and drop temporary revenue column}
df_final <- df1 %>% 
  select(-temp_revenue) %>% 
  mutate(advert_period = ifelse(sales_date < "2021-08-13", "before", "after" ))
```

```{r Reorder advertising period so that before comes first and then after}

df_final$advert_period <- factor(df_final$advert_period, levels=c('before','after'))
```



```{r Check dataset}
glimpse(df_final)
```


This dataset is 100% synthetic. It's simulated data, but let's visualize the data to make sure the dataset is good enough for performing our analysis.

## 3. Visualize dataset

```{r Let us visualize dataset}
p1 <- ggplot(df_final, aes(sales_date, revenue)) + geom_line() + theme_hc() + scale_y_continuous(expand = c(0, 0), limits = c(0, 7000)) + labs(x=" ", y=" ")
p2 <- ggplot(df_final, aes(revenue, advert_period)) + geom_boxplot() + theme_hc() + coord_flip() + labs(x = " ", y="Advertising Period")

plot_grid(p1, p2, nrow = 2, ncol = 1, labels = c("Revenue over time", "Revenue by social advertising period"))
```


## 4. Analysis


Now, let's create a new dataframe. We want equal number of weeks of before and during our social ad spending time periods. We spent during a 10 week period so we want the 10 weeks before.

```{r Create subset dataframe to the time period we are interested in}
#Filter on dates from the start date of our spending minus 10 weeks plus the entire social spending period. To subtract 10 weeks from our spending date, we will multiply 7 days by 10 to get the right interval in weeks.
before_after_df <- df_final %>% 
        filter(between(sales_date, as.Date('2021-08-13') - 7*10, as.Date('2021-10-15')))


before_after_df
```


Then we will take the average revenue for both time periods.

```{r Get mean revenue of the two time periods before and after}
ads_group <- before_after_df %>% 
  group_by(advert_period) %>% 
  summarize_at('revenue', .funs=mean) %>% 
  arrange(advert_period)
ads_group


ads_group
```
Let's extract the average revenue for each period and calculate the percentage difference. With the average difference in revenue, we can calculate the uplift.

```{r Before and after means}
beforeMean<- as.numeric(ads_group %>% filter(advert_period == 'before') %>% select(revenue))


afterMean <- as.numeric(ads_group %>% filter(advert_period == 'after') %>% select(revenue))
 
uplift <- round(afterMean - beforeMean,2) 


percChange<- round(((afterMean-beforeMean)/beforeMean) *100,2)


print(paste("The difference in revenue from before and after is", uplift, "US Dollars", "with the percentage difference being", percChange, "%"))
```

First, let's grab the average spend.


```{r Average Marketing Spend}
#First calculate the average spend
avg_spend_df <- df_final %>% 
  filter(between(sales_date, as.Date('2021-08-13'), as.Date('2021-10-15'))) %>% 
  summarize(avg_spend = mean(social))

(avg_spend <- avg_spend_df[[1]])
```


Now calculate the return on ad spend based on uplift and average spending.

```{r Return on advertising spend}
#Take the uplift which is the difference in average revenue subtract the average spend. Then divide by the average spend.
roas <- round((uplift - avg_spend)/avg_spend,2)
roas
```

If our revenue is in dollars, then for every $1 spent, we earned $2.16 dollars in revenue.


## 5. Final Summary



Before and after analysis can help provide some context to evaluate initiatives that could not be A/B tested, but we still want to understand the impact.



## 6. References

[1] Chapman, C. and McDonnell Feit, E., (2015). R for marketing research and analytics. Cham: Springer, pp.47-59.


```{r Export original dataset to a csv file, include=FALSE}
#Write csv file
write.csv(df_final, "datasets/marketing_impact_data.csv")

#Write to an Excel file
#Using openxlsx package
write.xlsx(x = df_final, file = "datasets/marketing_impact_data.xlsx")
```

