---
title: "How to simulate sales data in R"
subtitle: "Stretch your coding and data analysis skills by creating your own datasets"
author: "Anita Owens"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
#TOC {
  color: #000000;
  font-family: Calibri;
  font-size: 16px; 
  border-color: #708090;
}
#header {
  color: #000000;
  background-color: #edf2f4;
  opacity: 1;
  font-family: Calibri;
  font-size: 20px;
}
body {
  color: #000000;
  font-family: Calibri;
  background-color: #ffffff;
  font-size: 18px;
}
pre {
  color: #000000;
  background-color: #ffffff;
}
</style>
```

## 1. Introduction

Why simulate data?

Have you ever spent an inordinate amount of time looking for the right
data set to try out an analytical technique, but you don't quite find
what you are looking for.

Well, why not just create your own dataset for maximum flexibility which
gives you a few advantages:

1.  To test your analysis to make sure it's working.
2.  To make changes to your dataset to see what happens and how your
    conclusions might change based on different parameters . (e.g. How
    sales are impacted by different levels of the price of our product).

This article will walk through how to create synthetic data in R.

We want to simulate weekly sales of a product at a store. It could be an
in-store or ecommerce site, it doesn't really matter.


```{r Set up environment and load packages}

# Install pacman if needed
if (!require("pacman")) install.packages("pacman")

# load packages
pacman::p_load(pacman,
  tidyverse, openxlsx, psych, ggcorrplot, corrplot, cowplot, openxlsx)
```

## 2. Simulate Data

First, we will create a skeleton data frame to hold our variables and
initialize with 0 values for all columns except the date column.

1.  Sales date (based on week level granularity)
2.  Unit sales
3.  Unit price
4.  Paid social ads
5.  Promotion - whether there was a promotion this week.

```{r Create skeleton data frame}
#Initialize empty data frame
df <- data.frame(sales_date = seq(as.Date('2019-01-01'), by = 'week', length.out=52),
                 unit_sales = 0, price = 0, social = 0, promotion = 0)

#view data frame
df
```

Always set the seed for reproducibility when simulating data especially if you want to re-create the same dataset later.


```{r Set seed for reproducibility}
#Set seed for reproducibility
set.seed(90210)
```

We want to add weeks with marketing promotions.

We will simulate with the binomial distribution. The notation of the binomial distribution is B(n,p), where n is the number of experiments or trials, and p is the probability of a success.


```{r Create promotion value}
#Create weeks where there were promotions running. 10% likelihood of a promotion
df$promotion <- rbinom(n=nrow(df), size = 1, p=0.1)
```

We want to only add Paid Social values for the range of dates between
July and September and again in December.

```{r We want to replace weeks where we had paid social ads running for the amount we spent}
df <- df %>% 
  mutate(social = replace(social, between(sales_date,as.Date('2019-07-02'),as.Date('2019-09-10')), 350)) %>% mutate(social = replace(social, between(sales_date,as.Date('2019-12-03'),as.Date('2019-12-24')), 200))
```

Generate the product's price and place in a vector and then we'll use the sample function to randomly add prices to our dataframe.

```{r Create price values}
df$price <- sample(x=c(4.50, 4.99), size = nrow(df), replace = TRUE)

#Check output
table(df$price)
```


Next is to generate sales data and place into a temporary sales
variable.

Sales data is randomly generated based on a poisson distribution.

```{r Generate Sales Data}
#first argument is the number of draws
#2nd argument is the mean value per week of sales
temp_sales <- rpois(nrow(df), lambda = 8300)

#Check output
temp_sales
```


```{r Scale sales up according to price to follow a logarithmic function}
temp_sales <- temp_sales * log(df$price)

#Check output
temp_sales
```



In our last step, we need to take our temporary sales and make sure to
add an increase of 30% in unit sales for weeks we have the promotion
running. We want to show the boost in sales in the weeks where there are promotions.

```{r Add impact of increased sales due to week where a promotion was running}
df$unit_sales <- floor(temp_sales * (1 + df$promotion * 0.30))
```

## 3. Check Results

```{r Descriptive statitics}
#Let's do some descriptive statistics
df %>% 
 select_if(is.numeric) %>% 
  psych::describe()
```

Let's visualize the variables in our dataset we created.

```{r Visualize Variables Individually}
#Create visualizations
p1 <- ggplot(df, aes(sales_date, unit_sales)) + geom_line() + theme_minimal() 

p2 <- ggplot(df, aes(social, unit_sales)) + geom_point() + theme_minimal() 

p3 <- ggplot(df, aes(factor(promotion), unit_sales)) + geom_boxplot() + theme_minimal()

p4 <- ggplot(df, aes(factor(price), unit_sales)) + geom_boxplot() + theme_minimal()


plot_grid(p1, p2, p3, p4, labels = "AUTO")
```

Visualization of the correlation coefficients.

```{r Pairs Plot}
#Correlation Plot
df %>% 
 select_if(is.numeric) %>% 
  cor() %>% 
    corrplot(type = "upper", addCoef.col = "black", diag=FALSE)
```

Our synthetic data looks good enough for us to do some data analysis
e.g. marketing mix modeling, measuring advertising effectiveness, etc.

## 4. Export simulated dataset to an external file

```{r Save simulated dataset as a csv file or excel file}

write.csv(df, "weekly_sales_data.csv")

#Using openxlsx package
write.xlsx(x = df, file = "weekly_sales_data.xlsx")
```

## 5. References

1.  Chapman, C. and McDonnell Feit, E., (2015). R for marketing research
    and analytics. Cham: Springer, pp.47-59, 162--191.
